---
title: "Estimation and inference with hdcce"
output: rmarkdown::html_vignette
bibliography: null
vignette: >
  %\VignetteIndexEntry{Estimation and inference with hdcce}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 6.5, fig.height = 3.6, fig.align = "center")
options(width = 88)
set.seed(20260810)
```

This vignette shows how to use **hdcce** on a panel with many regressors and
unobserved interactive fixed effects. It covers the three things the package
does: estimating the slope vector, constructing confidence intervals for
individual coefficients, and testing whether a covariate enters the model at
all when the specification is additive rather than linear.

The methods are those of Rücker, Vogt, Linton and Walsh (2025) and Rücker,
Vogt and Linton (2026); this document explains what the functions compute and
how to read their output, not the theory behind them.

```{r}
library(hdcce)
```

# 1. The model and the data layout

For units $i = 1, \dots, n$ and periods $t = 1, \dots, T$,

$$Y_{it} = X_{it}^\top \beta + \gamma_i^\top F_t + \varepsilon_{it},$$

where $X_{it} \in \mathbb{R}^p$, the $F_t \in \mathbb{R}^K$ are unobserved
factors and the $\gamma_i$ are unit-specific loadings. The dimension $p$ may
exceed $nT$, and $\beta$ is assumed sparse. Neither the factors nor the
loadings are estimated: they are removed by projecting on the space spanned by
the cross-sectional averages of the regressors.

**The panel must be sorted by unit.** Rows $(i-1)T + 1, \dots, iT$ of `x` and
`y` belong to unit $i$. This is not checked, and a panel sorted by period will
return numbers rather than an error.

The package ships two simulated panels produced by `generate_data`, with
$n = T = 20$ and $p = 61$:

```{r}
data("data_estimation")
data("data_inference")

obs_N <- 20
obs_T <- 20

dim(data_estimation$x)
length(data_estimation$y)
```

The true coefficient vector has ten non-zero entries: the first regressor, and
the first three regressors of each of three groups.

```{r}
p     <- ncol(data_estimation$x)
gsize <- (p - 1) / 3
beta_true <- c(1, rep(c(1, 1, 1, rep(0, gsize - 3)), 3))
which(beta_true != 0)
```

# 2. Estimation

`hdcce_estimator` performs the two steps of the procedure: it estimates the
factor space from the cross-sectional averages and projects the data on its
orthogonal complement, then runs a lasso on the projected data.

```{r}
fit <- hdcce_estimator(data_estimation, obs_N = obs_N, obs_T = obs_T,
                       NFOLDS = 5)
fit$K_hat
```

The number of factors is the count of normalised eigenvalues of
$\bar{\boldsymbol{X}}^\top \bar{\boldsymbol{X}} / T$ exceeding the truncation
`TRUNC`. The returned `eigenvalues` let you see how clear-cut that decision
was:

```{r scree}
plot(fit$eigenvalues[1:15], type = "b", pch = 16, ylim = c(0, 1),
     xlab = "index", ylab = "normalised eigenvalue",
     main = sprintf("K_hat = %d", fit$K_hat))
abline(h = 0.01, col = "red", lty = 2)
legend("topright", "TRUNC", lty = 2, col = "red", bty = "n")
```

A sharp drop after the third eigenvalue is what one hopes to see; a gradual
decay means the factor number is not well identified and the result will be
sensitive to `TRUNC`. Supplying `NFACTORS` fixes $\widehat{K}$ directly.

The estimated coefficients recover the sparsity pattern, with the usual lasso
shrinkage towards zero:

```{r}
round(head(as.numeric(fit$coefs), 6), 3)

est <- as.numeric(fit$coefs)
c(selected      = sum(est != 0),
  true_nonzero  = sum(beta_true != 0),
  found         = sum(est[beta_true != 0] != 0))
```

More coefficients are selected than are truly non-zero. That is expected when
the penalty is chosen by cross-validation, which targets prediction rather than
selection, and it is the reason inference requires the debiasing step below
rather than reading significance off the lasso fit.

# 3. Confidence intervals

`hdcce_inference` returns a desparsified estimate for each coefficient in
`COEF_INDEX_VEC`. The debiasing runs a nodewise lasso of the target regressor
on the others and corrects the lasso estimate along the resulting residual
direction, which restores asymptotic normality.

The second shipped panel has three response columns, differing only in the
coefficient of the first regressor: $c^{**} = 0$, $0.1$ and $0.2$.

```{r}
dat0 <- list(x = data_inference$x, y = data_inference$y[, 1])   # beta_1 = 0
inf0 <- hdcce_inference(dat0, obs_N = obs_N, obs_T = obs_T,
                        COEF_INDEX_VEC = 1, NFOLDS = 5)

r <- inf0$results[["1"]]
c(estimate = r$coef_despar, se = r$se, p_value = r$p_value)
r$confidence_band
```

Rows of `confidence_band` follow the order of `alpha`, which defaults to
`c(0.01, 0.05, 0.10)`. The interval covers zero, as it should: the truth here
is $\beta_1 = 0$.

With a non-zero coefficient the interval moves away from zero:

```{r}
dat2 <- list(x = data_inference$x, y = data_inference$y[, 3])   # beta_1 = 0.2
inf2 <- hdcce_inference(dat2, obs_N = obs_N, obs_T = obs_T,
                        COEF_INDEX_VEC = 1, NFOLDS = 5)
inf2$results[["1"]]$confidence_band
```

The reported $p$-value tests $H_0 \colon \beta_j = 0$ and is exactly dual to the
band: $p \le \alpha$ if and only if zero lies outside the $(1-\alpha)$ interval.

```{r}
alpha <- inf0$alpha
cb    <- r$confidence_band
cbind(alpha,
      p_le_alpha    = r$p_value <= alpha,
      zero_excluded = !(cb[, 1] <= 0 & 0 <= cb[, 2]))
```

Several coefficients can be treated in one call; the result carries one entry
per index, named by it.

```{r}
inf_multi <- hdcce_inference(dat0, obs_N = obs_N, obs_T = obs_T,
                             COEF_INDEX_VEC = c(1, 2, 5), NFOLDS = 5)
names(inf_multi$results)
t(sapply(inf_multi$results,
         function(z) c(estimate = z$coef_despar, se = z$se, p = z$p_value)))
```

Coefficients 1 and 5 are truly zero, coefficient 2 is truly one.

The `HAC` argument selects the variance estimator: `1` assumes homoscedastic
errors without serial correlation, `2` (the default) allows heteroscedasticity,
and `3` allows serial correlation as well. The choice matters in small $T$,
where the unit-wise variance estimates underlying `HAC = 2` rest on few degrees
of freedom.

```{r}
sapply(1:3, function(k)
  hdcce_inference(dat2, obs_N, obs_T, COEF_INDEX_VEC = 1,
                  NFOLDS = 5, HAC = k)$results[["1"]]$se)
```

# 4. Testing a covariate in an additive model

Supplying `dictionaries` replaces the linear index by

$$Y_{it} = \sum_{j=1}^p m_j(X_{it,j}) + \gamma_i^\top F_t + \varepsilon_{it},
\qquad m_j(x) = \phi_j(x)^\top \beta_j,$$

where $\phi_j$ is a known vector of $L_j$ transformations. The function then
tests $H_0 \colon m_j = 0$ against $H_1 \colon m_j \neq 0$, which is a
functional null: there is no scalar to invert, so the output is a test rather
than a confidence interval.

A dictionary is a matrix `Phi` of stacked transformations together with a
vector `group` recording which covariate each of its columns transforms. With
$\phi_j(x) = (x, x^2)$:

```{r}
X   <- data_inference$x
Phi <- do.call(cbind, lapply(1:p, function(j) cbind(X[, j], X[, j]^2)))
dict <- list(Phi = Phi, group = rep(1:p, each = 2))
dim(Phi)
```

We check, whether the first covariate has an influence at all:
```{r}
tst <- hdcce_inference(list(x = X, y = data_inference$y[, 1]),
                       obs_N = obs_N, obs_T = obs_T, COEF_INDEX_VEC = 1,
                       dictionaries = dict, NFOLDS = 5, B = 500)

q1<- tst$results[["1"]]
c(statistic = q1$statistic, p_value = q1$p_value)
q1$critical_values
```



We check, whether the second covariate has an influence at all:
```{r}
tst <- hdcce_inference(list(x = X, y = data_inference$y[, 1]),
                       obs_N = obs_N, obs_T = obs_T, COEF_INDEX_VEC = 2,
                       dictionaries = dict, NFOLDS = 5, B = 500)

q2 <- tst$results[["2"]]
c(statistic = q2$statistic, p_value = q2$p_value)
q2$critical_values
```

The statistic is a maximum of self-normalised correlations between the
residuals under the null and kernel weights centred at a grid of locations
$w$; its critical values come from a Gaussian coupling, simulated `B` times.
Because it is a maximum, it is informative to look at where the evidence sits:

```{r profile1}
prof1 <- q1$profile
plot(prof1$w, abs(prof1$Psi_w), type = "b", pch = 16,
     xlab = "location w", ylab = expression(group("|", Psi[list(w,h)], "|")),
     ylim = range(0, abs(prof1$Psi_w), q1$critical_values))
abline(h = q1$critical_values[2], col = "red", lty = 2)
legend("topright", "5% critical value", lty = 2, col = "red", bty = "n")
```

for the first test and

```{r profile2}
prof2 <- q2$profile
plot(prof2$w, abs(prof2$Psi_w), type = "b", pch = 16,
     xlab = "location w", ylab = expression(group("|", Psi[list(w,h)], "|")),
     ylim = range(0, abs(prof2$Psi_w), q2$critical_values))
abline(h = q2$critical_values[2], col = "red", lty = 2)
legend("topright", "5% critical value", lty = 2, col = "red", bty = "n")
```


for the second test.


The `profile` also reports `n_eff`, the number of nodewise residuals falling in
each bump. This is the diagnostic to check before trusting the test: when the
smallest count is very low the Gaussian approximation is unreliable, and the
bandwidth should be increased or the region narrowed.


```{r}
prof1
```

and

```{r}
prof2
```


# 5. Two things worth checking

**The factor number.** Everything downstream depends on
$\widehat{\boldsymbol{\Pi}}$ removing the factors, and that in turn requires the
cross-sectional averages to identify the factor space — the usual CCE rank
condition, which asks the mean loading matrix to have rank $K$. When it fails,
$\widehat{K}$ is driven below the truth, a factor component survives in the
residuals, the estimated error variance inflates and the procedures lose power.
The scree plot above is the quick diagnostic.

**The occupancy of the grid.** For the test, `min(profile$n_eff)` says how much
data supports the sparsest direction. Compare it against the total (Test 1):

```{r}
c(min_n_eff = min(prof1$n_eff), max_n_eff = max(prof1$n_eff), nT = obs_N * obs_T)
```

# References

Rücker, M., Vogt, M., Linton, O. and Walsh, C. (2025). Estimation and inference
in high-dimensional panel data models with interactive fixed effects.
*Quantitative Economics* **16**(4), 1457–1509.
[doi:10.3982/QE2308](https://doi.org/10.3982/QE2308)

Rücker, M., Vogt, M. and Linton, O. (2026). High-dimensional panel data models
with interactive fixed effects: beyond the linear case.
[arXiv:2608.02055](https://arxiv.org/abs/2608.02055)
